---
title: "Spatial Models and Areal Unit Observations For Demographic Health Data"
author: Neal Marquez
date: October 31st, 2018
output: ioslides_presentation
---

# Background

## The Demand for High Resolution Geospatial Data

- National Averages Mask Inequalities
- Arbitrary Areal Units Can Do the Same
- Geospatial Inequalities Highlight Known Spatial Clusters of Other Clusters
    - SocioEconomic
    - Demographic
    - Cultural

## Data Limitations

- Data often does not come to us geo-referenced (Point)
- Methods have been developed for Areal and Point Data but not both
- Unclear on how we move between these two spaces

## Current Approaches

- Redistribution of Data Across Polygon of Source
- Joint Model Using Areal methods(LCAR) alongside Point methods(SPDE)
- Only use data of one type

## Problems
- Redistribution of Data Across Polygon of Source
    - Susceptible to spurious covariate correlations
    - Over-Confidence in redistributed areas
- Joint Model Using Areal methods(LCAR) alongside Point methods(SPDE)
    - Does not reflect the data generation process
    - Covariate interpretation susceptible to MAU problem
- Only use data of one type
    - Ignores information

# Proposal: Modification of Spatial Model Likelihood for Areal Response Data

## Traditional Continuous Spatial Model

- The Data Generating process
$$
\boldsymbol{Y} \sim \mathcal{GP}(\mu_\text{Constant}(\beta_0), K_{\text{Matern}}(\kappa, \sigma_\eta, \nu) + K_{\text{White}}(\sigma_\epsilon)) \\
$$

- GMRF Approximation
$$
i,j \in \{1, \dots, n\}\\
Y_i \sim \mathcal{N}(\beta_0 + \eta(s_i), \sigma_\epsilon) \\
\boldsymbol{\eta} \sim \text{MVN}(\boldsymbol{0}, \boldsymbol{\Sigma}) \\
\Sigma_{ij} = \frac{\sigma^2_\eta}{2^{\nu-1} \Gamma(\nu)}
    (\kappa ||s_i - s_j||)^\nu K_\nu(\kappa ||s_i - s_j||) \\
\Sigma_{ij} = \text{Cov}(\eta(s_i) , \eta(s_j)) \\
$$

## Extension to Areal Units  
$$
k \in \{ 1, \dots , m \} \\
l \in \mathbb{R} \\
Y_k \sim \mathcal{N}(\hat{y}_k, \sqrt{\sigma_\epsilon^2 + f(\langle \langle \mathcal{A}_k \rangle \rangle, \kappa, \sigma_\eta, \nu)}) \\
\hat{y}_k = \beta_0 + \frac{\int_{\forall l \in \mathcal{A}_k} \eta(s_l)ds}{\langle \langle \mathcal{A}_k \rangle \rangle} \\
$$

## Non Linear Binomial Likelihood of Points in Space
$$
i,j \in \{1, \dots, n\}\\
Y_i \sim \text{Binomial}(N_i, \hat{p}_i) \\
\text{logit}(\hat{p}_i) = X_i \boldsymbol{\beta} + \eta(s_i)  \\
\boldsymbol{\eta} \sim \text{MVN}(\boldsymbol{0}, \boldsymbol{\Sigma}) \\
\Sigma_{ij} = \frac{\sigma^2_\eta}{2^{\nu-1} \Gamma(\nu)}
    (\kappa ||s_i - s_j||)^\nu K_\nu(\kappa ||s_i - s_j||) \\
\Sigma_{ij} = \text{Cov}(\eta(s_i) , \eta(s_j))
$$

## Predictions of Evenly Distributed Population Areal Unit
$$
k \in \{ 1, \dots , m \} \\
l \in \mathbb{R} \\
Y_k \sim \text{Binomial}(N_k, \hat{p}_k) \\
\hat{p}_k = \frac{\int_{\forall l \in \mathcal{A}_k} \hat{p}_lds}{\langle \langle \mathcal{A}_k \rangle \rangle} \\
\hat{p}_k \approx \frac{\sum_l \hat{p}_l \Delta_l}{\langle \langle \mathcal{A}_k \rangle \rangle}=\frac{\sum_l \text{inv.logit}\big(X_l \boldsymbol{\beta} + \eta(s_l) \big) \Delta_l}{\langle \langle \mathcal{A}_k \rangle \rangle}
$$

# Simulation

## Data Generation

- Simulate Matern $\mathcal{GP}$ across field
- Add a constant $\beta_0 = -1$ to all points in space and take inverse logit
- Simulate data randomly from the field (N=1200, M=Poisson(100))
- Take representative data from random polygons and report outcome
    - Do not disclose exact location only polygons
- Estimate Field Using  
    - Traditional SPDE
    - Mix of Points and Polygons ussing the above approach
    - Polygons using the above approach

## Probability Field
```{r, pfield, echo=FALSE}
plotList <- readRDS("./simPlots.Rds")
plotList$fieldPlot
```

## County Field
```{r, pcounty, echo=FALSE}
plotList$countyPlot
```

# Traditional SPDE

## Traditional Model Observed Points on Field

```{r, fieldObs, echo=FALSE}
plotList$obsPointPlot
```

## Traditional Model Observed Points on Mesh
```{r, meshObs, echo=FALSE}
plotList$meshPointPlot
```

## Estimated Latent Field Vs True
```{r, estSPDE, echo=FALSE}
plotList$pointPlot
```

# Approximation with Point and Polygon Mix

## Simulation Paradigm
- Take originally simulated data and associate data with polygons
- Select half of the polygons and remove the points from the data
- Simulate data at random from removed polygons equal to points removed
    - Musr be representative of the polygon that was removed hence resampling

## New Data Structure
```{r, pointpolymixdata, echo=F}
plotList$ppDataMixPlot
```

## Projection Code

```
vector<Type> projPoint = AprojPoint * z;
vector<Type> projLatObs = AprojObs * z + beta0;
vector<Type> projPObs = exp(projLatObs) / (Type(1.) + exp(projLatObs));
SparseMatrix<Type> RAprojPoly = AprojPoly.transpose();
vector<Type> projPoly = RAprojPoly * projPObs;
```

## Mix Model Results
```{r, mixrez, echo=F}
plotList$mixPlot
```

# Polygon Only Model

## Results
```{r, poly, echo=F}
plotList$polyPlot
```

## Uncertainty in Latent Field
```{r, uncert, echo=F}
plotList$sdPlot
```

## Model Diagnostics

|Model  |Correlation|   RMSE| Coverage|
|:------|----------:|------:|--------:|
|Points |     0.9339| 0.0256|   0.9773|
|Mix    |     0.8668| 0.0345|   0.9502|
|Poly   |     0.7704| 0.0475|   0.9297|